---
title: "Getting started with cellpacman"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with cellpacman}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5, out.width = "100%",
  eval = TRUE
)
```

```{css, echo = FALSE}
.main-container { max-width: 1300px !important; }
img { max-width: 100% !important; }
```

## Overview

`cellpacman` (PAinted Cell and coMpound ANalysis) models how cell morphology
changes as the concentration of a compound increases, using well-level Cell
Painting features. For every *known* compound, i.e. one tested over a
concentration series, the package summarizes the morphological response as a
one-dimensional curve in a low-dimensional embedding: a **concentration
trajectory**. *Unknown* compounds, tested at a single concentration without
annotation, are then interpreted by projecting them onto these trajectories,
which yields both the most similar known compound and an interpolated
effective concentration.

The workflow is a chain of six functions. Each returns an S3 object that is
the input of the next, so intermediate results can be inspected, plotted, or
re-used:

| Step | Function | Output class | What it does |
| --- | --- | --- | --- |
| 1 | `loadCellPainting()` | `cellpainting_data` | Reads and validates the feature and metadata tables. |
| 2 | `normalize()` | `cellpainting_normalized` | z-scores every feature within plate. |
| 3 | `dimReduce()` | `cellpainting_dimred` | Embeds the normalized profiles in two dimensions with UMAP. |
| 4 | `compoundCluster()` | `cellpainting_clusters` | Runs DBSCAN per known compound and jointly on the unknown wells. |
| 5 | `curveEstimate()` | `cellpainting_curves` | Fits a principal curve through each known compound's concentration-ordered centroids. |
| 6 | `curveProject()` | list | Projects unknown-cluster centroids onto every trajectory and interpolates a concentration. |

Two feature-selection functions (`selectFeatureCurve()`,
`selectFeatureCluster()`) and two pairwise comparison functions
(`compareCompoundFeatures()`, `compareClusterFeatures()`) then explain *which*
features drive a trajectory or separate two groups of wells. This vignette
runs the complete workflow on the example screen bundled with the package.

```{r libraries}
library(cellpacman)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Input data

`cellpacman` expects two tables that share a `WellId` key:

* a **feature table** whose first column is `WellId` and whose remaining
  columns are numeric morphological features (one row per well), and
* a **metadata table** with at least `WellId`, `Plate`, `Compound`, and
  `Concentration`. Any additional columns are carried along unchanged.

`Compound` encodes the known/unknown distinction: a non-missing value marks a
well of a known compound, and `NA` marks an unknown well. `Concentration` must
be numeric and is only used for known compounds. Both tables may be supplied as
data frames or as paths to `csv`, `tsv`, `txt`, or `xlsx` files.

### The example screen

The package ships a Cell Painting screen of 1,920 wells from five 384-well
plates with 664 well-level features (`exampleDataPath()`). Each plate contains
three annotated compounds---axitinib (`Axit`), cabozantinib (`Cabo`), and
dactinomycin (`Dacti`)---at eight concentrations in duplicate, 16 wells of the
DMSO vehicle control, and 320 unannotated wells that enter the analysis as
unknown compounds.

```{r load}
dat <- loadCellPainting(
  file.feature  = exampleDataPath("features"),
  file.metadata = exampleDataPath("metadata")
)

dim(dat$feature_data)
dat$metadata |>
  mutate(Compound = ifelse(is.na(Compound), "(unknown)", Compound)) |>
  group_by(Compound) |>
  summarise(
    wells = n(),
    plates = n_distinct(Plate),
    concentrations = n_distinct(Concentration, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable()
```

### Preparing your own input

Instrument software usually exports one wide table per plate that mixes
features and annotation. The example tables were produced from such an export
by splitting on the column names; feature columns from Harmony all contain
`"NonBorder Cells"`. `readCellPainting()` preserves column names that contain
spaces, so the same split works on a fresh export:

```{r split, eval = FALSE}
raw <- readCellPainting("PlateResults.tsv")

feature_tbl  <- raw |> select(WellId, contains("NonBorder Cells"))
metadata_tbl <- raw |> select(!contains("NonBorder Cells"))

dat <- loadCellPainting(feature_tbl, metadata_tbl)
```

`loadCellPainting()` rejects duplicated `WellId`s, non-numeric feature
columns, missing required metadata columns, and feature wells without
metadata, so malformed input fails at this step rather than deep inside the
analysis.

## Preprocessing

### Within-plate standardization

Raw morphological features are measured on incommensurable scales and are
subject to plate-level technical variation. `normalize()` therefore
standardizes every feature within each plate: for feature $j$ on plate $k$,
each well value $x_{ijk}$ is replaced by
$z_{ijk} = (x_{ijk} - \bar{x}_{jk}) / s_{jk}$, where $\bar{x}_{jk}$ and
$s_{jk}$ are the mean and standard deviation over all wells of that plate.
Features with zero variance on a plate are set to 0. The `by` argument accepts
any set of metadata columns, so standardization can also be stratified by
batch or cell line.

```{r normalize}
dat_norm <- normalize(dat, method = "z.score", by = "Plate")
```

Standardizing against all wells on a plate (rather than against the vehicle
controls alone) assumes that most wells show a modest phenotype; this holds
for the typical screen in which unknown wells dominate the plate. The
normalized object `dat_norm` is reused by every feature-selection step below.

### Two-dimensional embedding

`dimReduce()` embeds the standardized profiles into two dimensions with UMAP
(`umap::umap()`, `n_components = 2`, `min_dist = 0.25`). The embedding is
deterministic for a given `random_seed`. Subsequent steps operate on these two
coordinates, stored as `x` and `y` in `reduced_data` together with the
metadata.

```{r dimreduce}
dat_dr <- dimReduce(dat_norm, method = "umap", random_seed = 1)
head(dat_dr$reduced_data[, c("WellId", "Plate", "Compound", "Concentration", "x", "y")])
```

`plotDimred()` draws the embedding with unknown wells in gray and known
compounds in color:

```{r plot-dimred, fig.height = 5}
plotDimred(dat_dr)
```

Colouring the same embedding by plate is a useful check that the within-plate
standardization has removed plate effects: the five plates should overlap
rather than form separate islands.

```{r plot-plate, fig.height = 5}
ggplot(dat_dr$reduced_data, aes(x, y, colour = Plate)) +
  geom_point(size = 0.6, alpha = 0.7) +
  labs(x = "DIM1", y = "DIM2", title = "UMAP embedding coloured by plate") +
  theme_bw()
```

## Clustering

Replicate wells of the same compound at the same concentration are expected to
occupy a common region of the embedding. `compoundCluster()` recovers these
groups with DBSCAN (`eps`, `minPts`), run

* **separately for each known compound**, using only that compound's wells, and
* **jointly for all unknown wells**.

DBSCAN labels points that do not belong to any dense region as noise (cluster
`"0"`). For each cluster the function records its centroid and, for known
compounds, the mean concentration of its members; known-compound clusters are
sorted by this mean concentration, which is what orders them along the
trajectory in the next step.

```{r cluster}
clusters <- compoundCluster(dat_dr, method = "dbscan", eps = 0.5, minPts = 5)

sapply(clusters$known_clusters, nrow)   # clusters per known compound
clusters$unknown_centroids              # unknown clusters
```

For the unknown wells, clustering is the primary result: each cluster is a
candidate phenotype that will be matched to a known compound.

```{r plot-unknown-clusters, fig.height = 5}
plotUnknownClusters(clusters)
```

For a known compound, the clusters summarize the concentration series. The
centroids (diamonds) are labelled with the mean concentration of their wells:

```{r plot-known-clusters, fig.height = 5}
plotKnownClusters(clusters, compound = "Cabo")
```

`eps` controls how finely a concentration series is resolved. A compound whose
wells form a single connected region at a given `eps` yields only one cluster
and therefore no trajectory; decreasing `eps` resolves sub-clusters, at the
cost of labelling more wells as noise. With `eps = 0.5` the dactinomycin wells
form one cluster, so no curve is fitted for `Dacti` below.

## Concentration trajectories

`curveEstimate()` summarizes the graded morphological response of each known
compound by a **principal curve** (Hastie and Stuetzle, 1989) fitted to that compound's
non-noise wells (`princurve::principal_curve()`, `smoother = "lowess"` by
default). The fit is initialized at the polyline that connects the cluster
centroids in order of increasing mean concentration, so the curve runs from
the low-concentration phenotype to the high-concentration phenotype. The arc
length $\lambda$ along the fitted curve is a pseudo-concentration axis.

Each centroid is mapped to the nearest point of the curve, which gives a set of
reference pairs $(\lambda_r, c_r)$ of arc length and mean concentration. These
pairs are the calibration that turns a position on the curve into an
estimated concentration. A curve requires at least two clusters; compounds
with fewer are skipped.

```{r curves}
curves <- curveEstimate(dat_dr, method = "princurve", eps = 0.5, minPts = 5, smoother = "lowess")

names(curves$curves)
curves$curves[["Cabo"]]$ref_points
```

The remainder of this vignette focuses on cabozantinib (`"Cabo"`).

```{r plot-known-curve, fig.width = 8, fig.height = 3.5}
focus_compound <- "Cabo"
plotKnownCurves(curves, compound = focus_compound)
```

## Matching unknown compounds to trajectories

`curveProject()` projects the centroid of every unknown cluster onto every
fitted trajectory (`princurve::project_to_curve()`). For each pair it reports

* `distance`: the Euclidean distance between the centroid and its projection
  (how far the unknown phenotype is from the trajectory);
* `arc_length`: the position of the projection along the curve; and
* `estimated_concentration`: the concentration obtained by piecewise-linear
  interpolation of the reference pairs at that arc length (positions beyond
  the first or last reference are clamped to the nearest reference).

`best_matches` keeps, for every unknown cluster, the trajectory with the
smallest distance.

```{r project}
proj <- curveProject(curves)

proj$best_matches |>
  transmute(
    `Unknown cluster` = unknown_cluster,
    `Best-match compound` = compound,
    `Estimated concentration` = signif(estimated_concentration, 3),
    `Distance to curve` = round(distance, 2)
  ) |>
  arrange(`Unknown cluster`) |>
  knitr::kable()
```

The full `projections` table holds every cluster-by-compound pair, so the
margin between the best and second-best match can be examined before a
cluster is interpreted. `plotUnknownProjections()` overlays the unknown
centroids, their projections, and the estimated concentrations on one
compound's trajectory:

```{r plot-projection, fig.width = 9, fig.height = 7}
plotUnknownProjections(curves, proj, compound = focus_compound)
```

## Feature selection

All feature-selection functions take the normalized data (`dat_norm`) and use
a two-sided Wilcoxon rank-sum test per feature, the difference in medians as
the effect size, and a multiplicity correction across features. The tests are
rank-based, so they are insensitive to the heavy tails that are common in
morphological features.

### Features that change along a trajectory

`selectFeatureCurve()` splits each known compound's wells at the median arc
length into a low- and a high-concentration group and tests every feature
between them (Bonferroni-adjusted by default). To show *how* a feature changes
rather than only *whether* it changes, the wells are also binned along the
arc length (`n_bins`) and the ten most significant features are averaged per
bin (`heatmap_data`).

```{r feature-curve, fig.width = 9, fig.height = 4}
feat_curve <- selectFeatureCurve(curves, data.norm = dat_norm, p.adjust = "bonferroni", cutoff = 0.05)

feat_curve[[focus_compound]]$top_10 |>
  mutate(feature = sub("^NonBorder Cells - (CP_ )?", "", feature)) |>
  select(feature, median_difference, adjusted_p_value, significance) |>
  knitr::kable(digits = 3)

feat_curve[[focus_compound]]$heatmap_data |>
  pivot_longer(-c(bin, mean_concentration), names_to = "feature", values_to = "value") |>
  mutate(feature = sub("^NonBorder Cells - (CP_ )?", "", feature)) |>
  ggplot(aes(factor(bin), feature, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b") +
  labs(
    x = "Trajectory bin (low to high concentration)", y = NULL, fill = "Mean z",
    title = paste("Top features along the", focus_compound, "trajectory")
  ) +
  theme_minimal(base_size = 9)
```

### Features that distinguish unknown clusters

`selectFeatureCluster()` compares every pair of unknown clusters and returns
one row per comparison and feature. The volcano plots show which features
separate the candidate phenotypes:

```{r feature-cluster, fig.width = 9, fig.height = 4.5}
feat_cluster <- selectFeatureCluster(clusters, data.norm = dat_norm)

ggplot(feat_cluster, aes(median_difference, -log10(adjusted_p_value), colour = significance)) +
  geom_point(size = 0.8, alpha = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed") +
  facet_wrap(~ comparison) +
  labs(x = "Median difference (second vs first cluster)", y = expression(-log[10]~"adjusted p")) +
  theme_minimal()
```

### Comparing any two groups

`compareCompoundFeatures()` contrasts any two compounds, where `NA` stands for
the pooled unknown wells, and `compareClusterFeatures()` contrasts two
clusters of one compound. Both use Benjamini--Hochberg adjustment by default.
Here the highest-concentration phenotype of cabozantinib is compared with the
vehicle control:

```{r compare}
compareCompoundFeatures(dat_norm, first = "DMSO", second = "Cabo") |>
  head(5) |>
  mutate(feature = sub("^NonBorder Cells - (CP_ )?", "", feature)) |>
  knitr::kable(digits = 3)
```

## Interactive analysis

The same workflow is available as a Shiny application, `cellpacman-shiny`,
which is distributed alongside the package and calls the functions used in
this vignette. It is described in `vignette("shiny-app", package = "cellpacman")`.

## References

Hastie, T. and Stuetzle, W. (1989). Principal curves. *Journal of the
American Statistical Association*, 84(406), 502--516.

Ester, M., Kriegel, H.-P., Sander, J. and Xu, X. (1996). A density-based
algorithm for discovering clusters in large spatial databases with noise.
*Proceedings of KDD-96*, 226--231.

McInnes, L., Healy, J. and Melville, J. (2018). UMAP: Uniform Manifold
Approximation and Projection for dimension reduction. *arXiv:1802.03426*.

## Session info

```{r session}
sessionInfo()
```
